\bf Calendar-based graphics for visualizing people's daily schedules

### 2019-02-18 main.Rmd
blinded version produced
minor tweaks based on reviews

abstract: |
  Calendars are broadly used in society to display temporal information, and events. This paper describes a new R package with functionality to organize and display temporal data, collected on sub-daily resolution, into a calendar layout. The function `frame_calendar` uses linear algebra on the date variable to restructure data into a format lending itself to calendar layouts. The user can apply the grammar of graphics to create plots inside each calendar cell, and thus the displays synchronize neatly with **ggplot2** graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package **sugrrants**.

@@

We develop a method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout. The calendar format is created using linear algebra, giving a restructuring of the data, that can then be piped into grammar of graphics definitions of plots, as used in **ggplot2** [@R-ggplot2]. The data restructuring approach is consistent with the tidy data principles available in the **tidyverse** [@R-tidyverse] suite. The methods are implemented in a new package called **sugrrants** [@R-sugrrants] .

The purpose of the calendar-based visualization is to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia [@ped]. There have been 43 sensors installed that count pedestrians every hour across the inner-city area till the end of 2016 (Figure \ref{fig:ped-map}). The data set can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. A routine examination of the data would involve constructing conventional time series plots to catch a glimpse of temporal patterns. The faceted plots in Figure \ref{fig:time-series-plot}, give an overall picture of the foot traffic at three different sensors over 2016. Further faceting by day of the week (Figure \ref{fig:facet-time}) provides a better glimpse of the daily and sub-daily pedestrian patterns.

<!--
1. Variations exist on multiple time scales including time of day, day of week, and day of year (such as public holiday and recurring events).
2. Since the data are often collected at sub-daily frequencies, they typically involve a large number of observations.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need for comparing and contrasting between locations.
-->

@@

```{r ped-map, fig.width = 7, out.width = "0.7\\linewidth", fig.cap = "Map of the Melbourne city area with dots indicating sensor locations. These three highlighted sensors will be inspected in the paper: (1) the State Library--a public library, (2) Flagstaff Station--a train station, closed on non-work days, (3) Birrarung Marr: an outdoor park hosting many cultural and sports events."}
```

@@

The work is inspired by @Wickham2012glyph, which uses linear algebra to display spatio-temporal data as glyphs on maps.  It is also related to recent work by @R-geofacet which provides methods in the **geofacet** package to arrange data plots into a grid, while preserving the geographical position. Both of these show data in a spatial context.

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities, and special events. There is some existing work in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: **ggTimeSeries** [@R-ggTimeSeries] and **ggcal** [@R-ggcal]. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

Figure \ref{fig:fs-2016} shows the line glyphs framed in the monthly calendar over the year 2016. This is achieved by the `frame_calendar` function, which computes the coordinates on the calendar for the input data variables. These can then be plotted using the usual **ggplot2** package [@R-ggplot2] functions. All of the grammar of graphics [@wilkinson2006grammar; @wickham2009ggplot2] can be applied.

@@

By default, grids are laid out horizontally. This can be transposed by setting the `dir` parameter to `"v"`, in which case $i$ and $j$ are swapped in the Equation \ref{eq:grid}. This can be useful for creating  calendar layouts for countries where vertical layout is the convention.

@@

Plots can be layered. The comparison of sensors can be done by overlaying plot the values for each (Figure \ref{fig:overlay}). Differences between the pedestrian patterns at these sensors can be seen. Flagstaff Station exhibits strong commuters patterns, with fewer pedestrian counts during the weekends and public holidays. This suggests that Flagstaff Station has limited functionality on non-work days. From Figure \ref{fig:overlay} it can be seen that Birrarung Marr has a distinct temporal trend from the other two all year round. The nighttime events, such as White Night, have barely affected the operation of Flagstaff Station but heavily affected the incoming and outgoing traffic to the State Library and Birrarung Marr.

@@

To avoid the overlapping problem, the calendar layout can be embedded into a series of subplots for the different sensors. Figure \ref{fig:facet} presents the idea of faceting calendar plots. This allows comparing the overall structure between sensors, while emphasizing individual sensor variation. In particular, it can be immediately learned that when Birrarung Marr was busy and packed, for example Australian Open in the last two weeks of January. This is concealed in the conventional graphics.

@@

The algorithm can also produce more complicated plots, such as boxplots. Figure \ref{fig:boxplot} uses a loess smooth line superimposed on side-by-side boxplots. It shows the distribution of hourly counts across all 43 sensors during December. The last week of December is the holiday season: people are off work on the day before Christmas, go shopping on the Boxing day, and stay out for the fireworks on New Year's Eve.

@@

As a data restructuring tool, the interactivity of calendar-based display can be easily enabled, as long as the interactive graphic system remains true to the spirit of the grammar of graphics, for example **plotly** [@plotly] in R. As a standalone display, an interactive tooltip can be added to show labels when mousing over it in the calendar layout, for example the hourly count with the time of day. It is difficult to sense the values from the static display, but the tooltip makes it possible. Options in function `frame_calendar` can be ported to a form of selection button or text input in a graphical user interface like R shiny [@R-shiny]. The display will update on the fly accordingly via clicking or text input, as desired.

@@

Figure \ref{fig:hod} shows energy consumption against time of day, separately by week day and weekend. Household 1 is an early bird, starting their day before 6 and going back home around 18 on week days. They switch air conditioning or heating on when they get home from work and keep it operating till mid-night, learned from the small horizontal cluster of points around 0.8 kWh. On the other hand, the stripes above 1 kWh for  household 2 indicates that perhaps air conditioning or heating runs continuously for some periods, consuming the twice the energy as household 1. A third peak occurs around 3pm for household 3 only, likely when the kids are home from school. They also have a consistent energy pattern between week days and weekends. As for household 4, their home routine starts after 18 on week days. Figures \ref{fig:dow} and \ref{fig:hod}, part of a traditional graphical toolkit, are useful for summarizing overall deviations across days and households.

@@

Figure \ref{fig:h1}, \ref{fig:h2}, \ref{fig:h3} and \ref{fig:h4} display the data in calendar layout individually for each household, unfolding their day-to-day life. Glancing over household 1, you can see that their overall energy use is low. Their week day energy use is distinguishable from their weekends, indicating a working household. The air conditioner appears to be used in the summer months in the evening and weekends. In household 2, heating keeps functioning for consecutive hours, which is evident in the mid July; while household 1 uses heating cautiously. These observations help to explain the stripes and clusters in Figure \ref{fig:hod}. The calendar plots speak the stories about vacation time that are untold by previous plots. Household 1 is on vacation over three weeks of June, and household 2 was also away for holidays during Christmas and the second week of June. Figure \ref{fig:h3} shows household 3 takes two one-month-long family trips in September till early October and in June/July, and household 4 is away over two or three weeks in early October, December and late June. The use of air conditioning and heating leaves no trace in these two households.

@@

The calendar layout will be useful for studying consumer trends and human behavior. It will not be so useful for physical patterns like climate, which are not typically affected by human activity. The layout does not replace traditional displays, but serves to complement to further tease out structure in temporal data. Analysts would still be advised to plot overall summaries and deviations, in order to study general trends.

The layout is a type of faceting and could be useful to develop this into a fully-fledged faceting method, with formal labels and axes. This is a future goal.

@@

We would like to thank Stuart Lee and Heike Hofmann for their feedback about this work. The most recent version of the `frame_calendar` function is included in the **sugrrants** package, which can be accessed via the CRAN website <https://CRAN.R-project.org/package=sugrrants> or Github <https://github.com/earowang/sugrrants>. All materials required to reproduce this article and a history of the changes can be found at the project's Github repository <https://github.com/earowang/paper-calendar-vis>.

```{r write-bib}
write_bib(c("ggplot2", "tidyverse", "sugrrants", "geofacet", "shiny", .packages()), 'packages.bib')
```

>>>>

abstract: |
  Calendars are broadly used in society to display temporal information and events. This paper describes a new R package with functionality to organize and display temporal data, collected on sub-daily resolution, into a calendar layout. The function `frame_calendar` uses linear algebra on the date variable to restructure data into a format lending itself to calendar layouts. The user can apply the grammar of graphics to create plots inside each calendar cell, and thus the displays synchronize neatly with **ggplot2** graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package **sugrrants**.

@@

We develop a method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout. The calendar format is created using linear algebra, giving a restructuring of the data, that can then be piped into grammar of graphics definitions of plots [@wilkinson2006grammar; @wickham2009ggplot2], as used in **ggplot2** [@R-ggplot2]. The data restructuring approach is consistent with the tidy data principles available in the **tidyverse** [@R-tidyverse] suite. The methods are implemented in a new R package called **sugrrants** [@R-sugrrants].

The purpose of the calendar-based visualization is to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia [@ped]. There have been 43 sensors installed that count pedestrians every hour across the inner-city area until the end of 2016 (Figure \ref{fig:ped-map}). The data set can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. A routine examination of the data would involve constructing conventional time series plots to catch a glimpse of temporal patterns. The faceted plots in Figure \ref{fig:time-series-plot} give an overall picture of the foot traffic at three different sensors over 2016. Further faceting by day of the week (Figure \ref{fig:facet-time}) provides a better glimpse of the daily and sub-daily pedestrian patterns.

@@

```{r ped-map, fig.width = 7, out.width = "0.7\\linewidth", fig.cap = "Map of the Melbourne city area with dots indicating sensor locations. These three highlighted sensors will be inspected in the paper: (1) the State Library---a public library, (2) Flagstaff Station---a train station, closed on non-work days, (3) Birrarung Marr---an outdoor park hosting many cultural and sports events."}
```

@@

The work is inspired by @Wickham2012glyph, which uses linear algebra to display spatio-temporal data as glyphs on maps.  It is also related to recent work by @R-geofacet which provides methods in the **geofacet** R package to arrange data plots into a grid, while preserving the geographical position. Both of these show data in a spatial context.

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There is some existing work in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: **ggTimeSeries** [@R-ggTimeSeries] and **ggcal** [@R-ggcal]. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

Figure \ref{fig:fs-2016} shows the line glyphs framed in the monthly calendar over the year 2016. This is achieved by the `frame_calendar` function, which computes the coordinates on the calendar for the input data variables. These can then be plotted using the usual **ggplot2** R package [@R-ggplot2] functions. All of the grammar of graphics can be applied.

@@

By default, grids are laid out horizontally. This can be transposed by setting the `dir` parameter to `"v"`, in which case $i$ and $j$ are swapped in Equation \ref{eq:grid}. This can be useful for creating  calendar layouts for countries where vertical layout is the convention.

@@

Plots can be layered. The comparison of sensors can be done by overlaying plot the values for each (Figure \ref{fig:overlay}). Differences between the pedestrian patterns at these sensors can be seen. Flagstaff Station exhibits strong commuters patterns, with fewer pedestrian counts during the weekends and public holidays. This suggests that Flagstaff Station has limited functionality on non-work days. From Figure \ref{fig:overlay} it can be seen that Birrarung Marr has a distinct temporal pattern from the other two all year round. The nighttime events, such as White Night, have barely affected the operation of Flagstaff Station but heavily affected the incoming and outgoing traffic to the State Library and Birrarung Marr.

@@

To avoid the overlapping problem, the calendar layout can be embedded into a series of subplots for the different sensors. Figure \ref{fig:facet} presents the idea of faceting calendar plots. This allows comparing the overall structure between sensors, while emphasizing individual sensor variation. In particular, it can be immediately learned that when Birrarung Marr was busy and packed, for example during the Australian Open, a major international tennis tournament, in the last two weeks of January. This is concealed in the conventional graphics.

@@

The algorithm can also produce more complicated plots, such as boxplots. Figure \ref{fig:boxplot} uses a loess smooth line superimposed on side-by-side boxplots. It shows the distribution of hourly counts across all 43 sensors during December. The last week of December is the holiday season: people are off work on the day before Christmas (December 24), go shopping on the Boxing day (December 26), and stay out for the fireworks on New Year's Eve.

@@

As a data restructuring tool, the interactivity of calendar-based displays can be easily enabled, as long as the interactive graphic system remains true to the spirit of the grammar of graphics, for example **plotly** [@plotly] in R. As a standalone display, an interactive tooltip can be added to show labels when mousing over it in the calendar layout, for example the hourly count with the time of day. It is difficult to sense the values from the static display, but the tooltip makes it possible. Options in the `frame_calendar` function can be ported to a form of selection button or text input in a graphical user interface like R shiny [@R-shiny]. The display will update on the fly accordingly via clicking or text input, as desired.

@@

Figure \ref{fig:hod} shows energy consumption against time of day, separately by week day and weekend. Household 1 is an early bird, starting their day before 6 and going back home around 18 on week days. They switch air conditioning or heating on when they get home from work and keep it operating until mid-night, learned from the small horizontal cluster of points around 0.8 kWh. On the other hand, the stripes above 1 kWh for  household 2 indicates that perhaps air conditioning or heating runs continuously for some periods, consuming the twice the energy as household 1. A third peak occurs around 3pm for household 3 only, likely when the kids are home from school. They also have a consistent energy pattern between week days and weekends. As for household 4, their home routine starts after 18 on week days. Figures \ref{fig:dow} and \ref{fig:hod}, part of a traditional graphical toolkit, are useful for summarizing overall deviations across days and households.

@@

Figure \ref{fig:h1}, \ref{fig:h2}, \ref{fig:h3} and \ref{fig:h4} display the data in calendar layout individually for each household, unfolding their day-to-day life. Glancing over household 1, we can see that their overall energy use is low. Their week day energy use is distinguishable from their weekends, indicating a working household. The air conditioner appears to be used in the summer months (December--February) in the evening and weekends. In household 2, heating keeps functioning for consecutive hours, which is evident in the mid July. In contrast, household 1 uses heating cautiously. These observations help to explain the stripes and clusters in Figure \ref{fig:hod}. The calendar plots speak the stories about vacation time that are untold by previous plots. Household 1 is on vacation over three weeks of June, and household 2 was also away for vacation in late December and in the second week of June. Figure \ref{fig:h3} shows household 3 takes two one-month-long family trips in September until early October and in June/July. Household 4 is away over two or three weeks in early October, December and late June. The use of air conditioning and heating leaves no trace in these two households.

@@

The calendar layout will be useful for studying consumer trends and human behavior. It will not be so useful for physical patterns like climate, which are not typically affected by human activity. The layout does not replace traditional displays, but serves to complement to further tease out structure in temporal data. Analysts would still be advised to plot overall summaries and deviations in order to study general trends.

The layout is a type of faceting and could be useful to develop this into a fully-fledged faceting method with formal labels and axes. This is a future goal.

@@

We would like to thank Stuart Lee and Heike Hofmann for their feedback about this work. The most recent version of the `frame_calendar` function is included in the **sugrrants** R package, which can be accessed via the CRAN website <https://CRAN.R-project.org/package=sugrrants> or Github <https://github.com/earowang/sugrrants>. All materials required to reproduce this article and a history of the changes can be found at the project's Github repository <https://github.com/earowang/paper-calendar-vis>.

### 2019-02-18 main.tex
blinded version produced
minor tweaks based on reviews

\bigskip
\begin{abstract}
Calendars are broadly used in society to display temporal information,
and events. This paper describes a new R package with functionality to
organize and display temporal data, collected on sub-daily resolution,
into a calendar layout. The function \texttt{frame\_calendar} uses
linear algebra on the date variable to restructure data into a format
lending itself to calendar layouts. The user can apply the grammar of
graphics to create plots inside each calendar cell, and thus the
displays synchronize neatly with \textbf{ggplot2} graphics. The
motivating application is studying pedestrian behavior in Melbourne,
Australia, based on counts which are captured at hourly intervals by
sensors scattered around the city. Faceting by the usual features such
as day and month, was insufficient to examine the behavior. Making
displays on a monthly calendar format helps to understand pedestrian
patterns relative to events such as work days, weekends, holidays, and
special events. The layout algorithm has several format options and
variations. It is implemented in the R package \textbf{sugrrants}.
\end{abstract}

@@

We develop a method for organizing and visualizing temporal data,
collected at sub-daily intervals, into a calendar layout. The calendar
format is created using linear algebra, giving a restructuring of the
data, that can then be piped into grammar of graphics definitions of
plots, as used in \textbf{ggplot2} \citep{R-ggplot2}. The data
restructuring approach is consistent with the tidy data principles
available in the \textbf{tidyverse} \citep{R-tidyverse} suite. The
methods are implemented in a new package called \textbf{sugrrants}
\citep{R-sugrrants} .

The purpose of the calendar-based visualization is to provide insights
into people's daily schedules relative to events such as work days,
weekends, holidays, and special events. This work was originally
motivated by studying foot traffic in the city of Melbourne, Australia
\citep{ped}. There have been 43 sensors installed that count pedestrians
every hour across the inner-city area till the end of 2016 (Figure
\ref{fig:ped-map}). The data set can shed light on people's daily
rhythms, and assist the city administration and local businesses with
event planning and operational management. A routine examination of the
data would involve constructing conventional time series plots to catch
a glimpse of temporal patterns. The faceted plots in Figure
\ref{fig:time-series-plot}, give an overall picture of the foot traffic
at three different sensors over 2016. Further faceting by day of the
week (Figure \ref{fig:facet-time}) provides a better glimpse of the
daily and sub-daily pedestrian patterns.

@@

\caption{Map of the Melbourne city area with dots indicating sensor locations. These three highlighted sensors will be inspected in the paper: (1) the State Library--a public library, (2) Flagstaff Station--a train station, closed on non-work days, (3) Birrarung Marr: an outdoor park hosting many cultural and sports events.}\label{fig:ped-map}
\end{figure}

@@

The work is inspired by \citet{Wickham2012glyph}, which uses linear
algebra to display spatio-temporal data as glyphs on maps. It is also
related to recent work by \citet{R-geofacet} which provides methods in
the \textbf{geofacet} package to arrange data plots into a grid, while
preserving the geographical position. Both of these show data in a
spatial context.

In contrast, calendar-based graphics unpack the temporal variable, at
different resolutions, to digest multiple seasonalities, and special
events. There is some existing work in this area. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colors indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasonal variation such as school vacations, all of which have
influence over the turn-outs in the office. The calendar-based heatmap
was implemented in two R packages: \textbf{ggTimeSeries}
\citep{R-ggTimeSeries} and \textbf{ggcal} \citep{R-ggcal}. However,
these techniques are limited to color-encoding graphics and are unable
to use time scales smaller than a day. Time of day, which serves as one
of the most important aspects in explaining substantial variations
arising from the pedestrian sensor data, will be neglected through daily
aggregation. Additionally, if simply using colored blocks rather than
curves, it may become perceptually difficult to estimate the shape
positions and changes, although using curves comes with the cost of more
display capacity \citep{cleveland1984graphical, lam2007overview}.

@@

Figure \ref{fig:fs-2016} shows the line glyphs framed in the monthly
calendar over the year 2016. This is achieved by the
\texttt{frame\_calendar} function, which computes the coordinates on the
calendar for the input data variables. These can then be plotted using
the usual \textbf{ggplot2} package \citep{R-ggplot2} functions. All of
the grammar of graphics \citep{wilkinson2006grammar, wickham2009ggplot2}
can be applied.

@@

By default, grids are laid out horizontally. This can be transposed by
setting the \texttt{dir} parameter to \texttt{"v"}, in which case \(i\)
and \(j\) are swapped in the Equation \ref{eq:grid}. This can be useful
for creating calendar layouts for countries where vertical layout is the
convention.

@@

As a data restructuring tool, the interactivity of calendar-based
display can be easily enabled, as long as the interactive graphic system
remains true to the spirit of the grammar of graphics, for example
\textbf{plotly} \citep{plotly} in R. As a standalone display, an
interactive tooltip can be added to show labels when mousing over it in
the calendar layout, for example the hourly count with the time of day.
It is difficult to sense the values from the static display, but the
tooltip makes it possible. Options in function \texttt{frame\_calendar}
can be ported to a form of selection button or text input in a graphical
user interface like R shiny \citep{R-shiny}. The display will update on
the fly accordingly via clicking or text input, as desired.

@@

We would like to thank Stuart Lee and Heike Hofmann for their feedback
about this work. The most recent version of the \texttt{frame\_calendar}
function is included in the \textbf{sugrrants} package, which can be
accessed via the CRAN website
\url{https://CRAN.R-project.org/package=sugrrants} or Github
\url{https://github.com/earowang/sugrrants}. All materials required to
reproduce this article and a history of the changes can be found at the
project's Github repository
\url{https://github.com/earowang/paper-calendar-vis}.

>>>>

\bigskip
\begin{abstract}
Calendars are broadly used in society to display temporal information
and events. This paper describes a new R package with functionality to
organize and display temporal data, collected on sub-daily resolution,
into a calendar layout. The function \texttt{frame\_calendar} uses
linear algebra on the date variable to restructure data into a format
lending itself to calendar layouts. The user can apply the grammar of
graphics to create plots inside each calendar cell, and thus the
displays synchronize neatly with \textbf{ggplot2} graphics. The
motivating application is studying pedestrian behavior in Melbourne,
Australia, based on counts which are captured at hourly intervals by
sensors scattered around the city. Faceting by the usual features such
as day and month, was insufficient to examine the behavior. Making
displays on a monthly calendar format helps to understand pedestrian
patterns relative to events such as work days, weekends, holidays, and
special events. The layout algorithm has several format options and
variations. It is implemented in the R package \textbf{sugrrants}.
\end{abstract}

@@

We develop a method for organizing and visualizing temporal data,
collected at sub-daily intervals, into a calendar layout. The calendar
format is created using linear algebra, giving a restructuring of the
data, that can then be piped into grammar of graphics definitions of
plots \citep{wilkinson2006grammar, wickham2009ggplot2}, as used in
\textbf{ggplot2} \citep{R-ggplot2}. The data restructuring approach is
consistent with the tidy data principles available in the
\textbf{tidyverse} \citep{R-tidyverse} suite. The methods are
implemented in a new R package called \textbf{sugrrants}
\citep{R-sugrrants}.

The purpose of the calendar-based visualization is to provide insights
into people's daily schedules relative to events such as work days,
weekends, holidays, and special events. This work was originally
motivated by studying foot traffic in the city of Melbourne, Australia
\citep{ped}. There have been 43 sensors installed that count pedestrians
every hour across the inner-city area until the end of 2016 (Figure
\ref{fig:ped-map}). The data set can shed light on people's daily
rhythms, and assist the city administration and local businesses with
event planning and operational management. A routine examination of the
data would involve constructing conventional time series plots to catch
a glimpse of temporal patterns. The faceted plots in Figure
\ref{fig:time-series-plot} give an overall picture of the foot traffic
at three different sensors over 2016. Further faceting by day of the
week (Figure \ref{fig:facet-time}) provides a better glimpse of the
daily and sub-daily pedestrian patterns.

@@

\caption{Map of the Melbourne city area with dots indicating sensor locations. These three highlighted sensors will be inspected in the paper: (1) the State Library---a public library, (2) Flagstaff Station---a train station, closed on non-work days, (3) Birrarung Marr---an outdoor park hosting many cultural and sports events.}\label{fig:ped-map}
\end{figure}

@@

The work is inspired by \citet{Wickham2012glyph}, which uses linear
algebra to display spatio-temporal data as glyphs on maps. It is also
related to recent work by \citet{R-geofacet} which provides methods in
the \textbf{geofacet} R package to arrange data plots into a grid, while
preserving the geographical position. Both of these show data in a
spatial context.

In contrast, calendar-based graphics unpack the temporal variable, at
different resolutions, to digest multiple seasonalities and special
events. There is some existing work in this area. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colors indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasonal variation such as school vacations, all of which have
influence over the turn-outs in the office. The calendar-based heatmap
was implemented in two R packages: \textbf{ggTimeSeries}
\citep{R-ggTimeSeries} and \textbf{ggcal} \citep{R-ggcal}. However,
these techniques are limited to color-encoding graphics and are unable
to use time scales smaller than a day. Time of day, which serves as one
of the most important aspects in explaining substantial variations
arising from the pedestrian sensor data, will be neglected through daily
aggregation. Additionally, if simply using colored blocks rather than
curves, it may become perceptually difficult to estimate the shape
positions and changes, although using curves comes with the cost of more
display capacity \citep{cleveland1984graphical, lam2007overview}.

@@

Figure \ref{fig:fs-2016} shows the line glyphs framed in the monthly
calendar over the year 2016. This is achieved by the
\texttt{frame\_calendar} function, which computes the coordinates on the
calendar for the input data variables. These can then be plotted using
the usual \textbf{ggplot2} R package \citep{R-ggplot2} functions. All of
the grammar of graphics can be applied.

@@

By default, grids are laid out horizontally. This can be transposed by
setting the \texttt{dir} parameter to \texttt{"v"}, in which case \(i\)
and \(j\) are swapped in Equation \ref{eq:grid}. This can be useful for
creating calendar layouts for countries where vertical layout is the
convention.

@@

As a data restructuring tool, the interactivity of calendar-based
displays can be easily enabled, as long as the interactive graphic
system remains true to the spirit of the grammar of graphics, for
example \textbf{plotly} \citep{plotly} in R. As a standalone display, an
interactive tooltip can be added to show labels when mousing over it in
the calendar layout, for example the hourly count with the time of day.
It is difficult to sense the values from the static display, but the
tooltip makes it possible. Options in the \texttt{frame\_calendar}
function can be ported to a form of selection button or text input in a
graphical user interface like R shiny \citep{R-shiny}. The display will
update on the fly accordingly via clicking or text input, as desired.

@@

We would like to thank Stuart Lee and Heike Hofmann for their feedback
about this work. The most recent version of the \texttt{frame\_calendar}
function is included in the \textbf{sugrrants} R package, which can be
accessed via the CRAN website
\url{https://CRAN.R-project.org/package=sugrrants} or Github
\url{https://github.com/earowang/sugrrants}. All materials required to
reproduce this article and a history of the changes can be found at the
project's Github repository
\url{https://github.com/earowang/paper-calendar-vis}.
